---
title: "Palms & Plots: Visualizing Trait Diversity in Palm Tree Dataset"
subtitle: "INFO 526 - Summer 2025 - Final Project"
author: 
  - name: "Aesthetic plotters"
    affiliations:
      - name: "School of Information, University of Arizona"
description: "his project aims to visualize how key biological traits shape the structural and ecological diversity of palm trees. We focus on two ecological questions, firstly, how fruit-related traits—size and visibility—influence the structural form of palms, measured via height; and secondly, whether palm defense traits—spines on leaves and stems—co-occur with growth form such as climbing versus erect habits."
format:
  html:
    code-tools: true
    code-overflow: wrap
    code-line-numbers: true
    embed-resources: true
editor: visual
code-annotations: hover
execute:
  warning: false
---

## **Package Setup**

```{r load_pkgs, message=FALSE, warning=FALSE, output=FALSE}
#| label: load-pkgs
#| message: false
if (!require(pacman))
  install.packages(pacman)

pacman::p_load(tidyverse,
               tidytuesdayR)
```

## Dataset

```{r}
#| label: load-dataset
#| message: false
# Data load using the tidytuesdayR package 
palm_tree_data <- tidytuesdayR::tt_load(2025, week = 11)

# Get all data in the dataset
palmtrees <- palm_tree_data$palmtrees

#Read raw data from csv file which is accessed from github repository
palmtrees <- readr::read_csv('data/palmtrees.csv')
```

The dataset we are using in this project is from the PalmTraits 1.0 database via the palmtrees R package by Emil Hvitfeldt. The dataset featured in March 18, 2025 TidyTuesday release and was originally published in the journel "PalmTraits 1.0, a species-level functional trait database of palms worldwide" by Kissling, W.D., Balslev, H., Baker, W.J. *et al* (2019). The dataset contains `r nrow(palmtrees)` rows and `r ncol(palmtrees)` columns. The dataset comes with a wide range of variables, the variables we will be using to analyse are :

```{r palm_variable_table, echo=FALSE}
#Load necessary libraries
library(tibble)
library(knitr)
# Create a summary table of key variables
tibble(
  Variable = c(
    "max_stem_height_m", 
    "fruit_size_categorical", 
    "conspicuousness", 
    "climbing", 
    "leaves_armed", 
    "stem_armed"
  ),
  Class = c(
    class(palmtrees$max_stem_height_m)[1],
    class(palmtrees$fruit_size_categorical)[1],
    class(palmtrees$conspicuousness)[1],
    class(palmtrees$climbing)[1],
    class(palmtrees$leaves_armed)[1],
    class(palmtrees$stem_armed)[1]
  ),
  Description = c(
    "Maximum stem height in meters",
    "Fruit size category: small (<4cm) or large (≥4cm)",
    "Fruit visibility: conspicuous (bright) or cryptic (dull)",
    "Growth habit: climbing or non-climbing",
    "Presence of spines on leaves",
    "Presence of spines on stem"
  )
) |> 
  # Adding table header
  kable(caption = "Key Variables to be used in this project")
```

### **Importance of choosing the dataset**

The database with its wide variety variables allows some rich comparative analysis and explanations. We have chosen this dataset as it exhibits some strong combination of biological significance, trait diversity, and visual capability. Its mix of categorical and continuous traits across thousands of species provides a meaningful opportunity to explore ecological patterns using effective data visualization. Additionally, its structure is well-suited to address research questions involving trait co-occurrence and plant strategy trade-offs—making it an ideal choice.

## Questions

The two questions you want to answer are:\
**Question 1**: Which palm traits exhibit the most variability across species, and how do fruit traits (size and conspicuousness) relate to stem height?\
**Question 2**: Do defense traits like spines on leaves and stems co-occur with particular growth habits (climbing vs erect palms)?

## Analysis plan

### **Approach for question 1**

Traits related to reproduction (fruit size and visibility) are expected to show moderate to low variation, high variation for structural growth traits, and palms with larger or more conspicuous fruits will tend to be taller, potentially due to selective pressure for enhanced seed dispersal.

For trait variability assessment we will create visual summaries with ridge plots for numeric traits and bar plots for categorical traits. To visualize the structural form of palm trees and how they are related to their fruit size and visibility, we will be using violin plots to visualize the distribution of maximum stem height (max_stem_height_m) across categories of fruit size and color conspicuousness. The conspicuousness variable represents how visually noticeable a fruit. Including it helps explore whether more visually striking fruits are associated with taller trees that may rely on long-distance seed dispersal. This will help us to examine the independent variables like fruit traits related to dependent variables like maximum stem height. Here the structural form is treated as the outcome variable, while fruit traits are predictors in our visual analysis.

Firstly, we start with filtering and cleaning the PlamTrait dataset, focus on key variables max_stem_height_m, fruit_size_categorical, and conspicuousness. We will remove any records with missing values and ensure categorical values are consistent.

The violin plot will display fruit_size_categorical on the x-axis and max_stem_height_m on the y-axis. We will fill the violins by the conspicuousness variable to add a second layer of categorical comparison. This will allow us to identify how palm height distributions differ across combinations of fruit size and visibility traits.

By using violin plots, we can observe the median heights and also the variability in height within each category. This design results in four combinations. We'll ensure clarity in the plot by either separating violins with facetting or introducing one variable at a time This approach helps us in visualizing whether taller palms are more likely to produce larger or more conspicuous fruits, and helps reveal potential evolutionary or ecological strategies linked to display and dispersal.

### **Approach for question 2**

Climbing palms are expected to exhibit fewer spine traits due to structural support from host plants, which may reduce herbivore exposure and the need for physical defenses.

For the second question we are using heatmap to visualize combination of spine traits and growth forms. This allows us to investiage defensive traits such as spines (armature) co-occur with certain growth habits (climbing vs non-climbing). The approach is ideal for categorical co-occurrence data.

We begin with filtering and cleaning of the PlamTrait dataset thereby retaining only the relevant variables that is climbing, leaves_armed, and stem_armed. Any missing or ambiguous entries will be excluded. We will standardize the categorical values for consistency (armed vs unarmed) and create a new variable representing each unique trait combination such as both armed, leaf only, stem only, unarmed.

We will then group the data by climbing status and the combination variable. Using dplyr, we’ll calculate the count of palm species in each category combination.\
\
The Heatmap will have "climbing" on one axis and the "spine combination categories" on the other. The fill color of each cell will reflect the "frequency or relative proportion of species" within that category pairing. We then annotate the tiles with exact counts to show visual interpretation better.\
\
Heat map visualization will help us identify whether climbing palms are more or less likely to exhibit defensive traits compared to erect palms. The heatmap’s clear structure makes it easy to detect trait co-occurrence patterns and generate hypotheses about the evolutionary or functional relationships between growth form and physical defenses.

\

## Plan of Attack

+----------------------------------------------+--------------------+-----------+------------+-------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| Task Name                                    | Assignee           | Status    | Due        | Priority          | Summary                                                                                                                                                                                                                                                                       |
+==============================================+====================+===========+============+===================+===============================================================================================================================================================================================================================================================================+
| Prepare proposal questions                   | Nilotpal & Sravani | Completed | 06-01-2025 | High Priority     | -   Analysing all the datasets from 2025 Github repository.                                                                                                                                                                                                                   |
|                                              |                    |           |            |                   |                                                                                                                                                                                                                                                                               |
|                                              |                    |           |            |                   | -   Make possible research questions out of those dataset.                                                                                                                                                                                                                    |
+----------------------------------------------+--------------------+-----------+------------+-------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| Finalize dataset & research questions        | Nilotpal & Sravani | Completed | 06-09-2025 | High Priority     | -   Selected the dataset [PalmTraits](https://www.nature.com/articles/s41597-019-0189-0)                                                                                                                                                                                      |
|                                              |                    |           |            |                   |                                                                                                                                                                                                                                                                               |
|                                              |                    |           |            |                   | -   Two research questions are finalized                                                                                                                                                                                                                                      |
+----------------------------------------------+--------------------+-----------+------------+-------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| Final Proposal Peer Review                   | Nilotpal & Sravani | Completed | 06-13-2025 | High Priority     | -   Sravani peer reviewed "A Statistical Exploration of Himalayan Expeditions".                                                                                                                                                                                               |
|                                              |                    |           |            |                   |                                                                                                                                                                                                                                                                               |
|                                              |                    |           |            |                   | -   Nilotpal peer reviewed "Crime in Arizona and Across the United States".                                                                                                                                                                                                   |
+----------------------------------------------+--------------------+-----------+------------+-------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| Edit proposal according to peer review       | Nilotpal & Sravani | Completed | 06-18-2025 | High Priority     | -   Changes in the proposal according to the peer review by [nathanielcross04](https://github.com/nathanielcross04) & [MattOsterhoudt](https://github.com/MattOsterhoudt), and also considered certain points that are to be taken care of as we work further in the project. |
+----------------------------------------------+--------------------+-----------+------------+-------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| Edit proposal according to instructor review | Nilotpal & Sravani | Completed | 06-23-2025 | High Priority     | -   Made changes to the questions and added hypotheses according to instructor review. Proposal file has been updated and all github issues created is moved to closed.                                                                                                       |
+----------------------------------------------+--------------------+-----------+------------+-------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| Cleaning the data, creating new variables    | Nilotpal & Sravani | To-Do     | 06-23-2025 | Moderate Priority |                                                                                                                                                                                                                                                                               |
+----------------------------------------------+--------------------+-----------+------------+-------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| Creating draft plots for question 1          | Sravani            | To-Do     | 06-24-2025 | High Priority     |                                                                                                                                                                                                                                                                               |
+----------------------------------------------+--------------------+-----------+------------+-------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| Creating draft plots for question 2          | Nilotpal           | To-Do     | 06-24-2025 | High Priority     |                                                                                                                                                                                                                                                                               |
+----------------------------------------------+--------------------+-----------+------------+-------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| Discussion on prepared graph                 | Nilotpal & Sravani | To-Do     | 06-25-2025 | Low Priority      |                                                                                                                                                                                                                                                                               |
+----------------------------------------------+--------------------+-----------+------------+-------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| Clean up code, Improve graph                 | Nilotpal & Sravani | To-Do     | 06-25-2025 | Moderate Priority |                                                                                                                                                                                                                                                                               |
+----------------------------------------------+--------------------+-----------+------------+-------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| Analyze and write-up visualization report    | Sravani            | To-Do     | 06-27-2025 | Moderate Priority |                                                                                                                                                                                                                                                                               |
+----------------------------------------------+--------------------+-----------+------------+-------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| Create PPT slides                            | Nilotpal           | To-Do     | 06-27-2025 | Moderate Priority |                                                                                                                                                                                                                                                                               |
+----------------------------------------------+--------------------+-----------+------------+-------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| Record presentation                          | Nilotpal & Sravani | To-Do     | 06-29-2025 | High Priority     |                                                                                                                                                                                                                                                                               |
+----------------------------------------------+--------------------+-----------+------------+-------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| Submit final presentation                    | Nilotpal & Sravani | To-Do     | 06-30-2025 | Low Priority      |                                                                                                                                                                                                                                                                               |
+----------------------------------------------+--------------------+-----------+------------+-------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| Final Project Submission                     | Nilotpal & Sravani | To-Do     | 07-02-2025 | Low Priority      |                                                                                                                                                                                                                                                                               |
+----------------------------------------------+--------------------+-----------+------------+-------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
